---
title: "LPI neutral theory"
author: "James G. Hagan"
date: "05/08/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r include=FALSE}
# load libaries
library(tidyverse)
library(here)
library(vegan)

# load the LPI data
lpi_dat_raw <- read_csv(file = here("data/LPI_LPR2016data_public.csv"), 
                        na = c("", "NULL"),
                        col_types = cols(`2015` = col_double()))

lpi_dat_raw

# check which variables are included
names(lpi_dat_raw)

# check for any parsing problems
problems(lpi_dat_raw)

# check the data structure
str(lpi_dat_raw)

# remove years before 1970 and after 2012 because there is insufficient data (McCrae et al. 2017)
lpi_dat_raw <- 
  lpi_dat_raw %>%
  select(!starts_with(c("195", "196", "2013", "2014", "2015")))

# GAMs are used to extroplate time-series with > 6 data points between start and end years
# i.e. they don't extrapolate beyond range of the data

# Chain method is used for time-series with < 6 data points
# log-linear imputation when year(-1) = X, year(0) = unknown, year(+1) = known
# the unknown year is then imputed

# this makes it complicated to determine how this is calculated
# specifically, which series are used in the calculation for each year


# determine how the population data has changed through time
# make a year and population size column

lpi_dat_raw %>%
  select(starts_with(c("19", "20"))) %>%
  names()

names(lpi_dat_raw)

lpi_pop <- 
  lpi_dat_raw %>%
  pivot_longer(cols = starts_with(c("19", "20")),
               names_to = "year",
               values_to = "population_size")


# how many unique population time-series are there?
n_series <- 
  lpi_pop$ID %>%
  unique(.) %>%
  length(.)

# how many unique population time-series have more than 6 data points?
n_series_6 <- 
  lpi_pop %>%
  filter(!is.na(population_size)) %>%
  group_by(ID) %>%
  summarise(n = n(), .groups = "drop") %>%
  filter(n > 6) %>%
  nrow()

n_series_6/n_series # <50% have more than 6 data points

# how many starting years are after 1970?
pop_id <- 
  lpi_pop %>%
  filter(!is.na(population_size)) %>%
  group_by(ID) %>%
  summarise(min_year = min(year),
            max_year = max(year),
            n = n(),
            .groups = "drop")


# create a variable with years between start and end date of each population
# 'Population time series with fewer than six data points or that resulted in poor GAM fit were modelled using the chain method [9]'
# from McCrae et al. (2017)

n0_nt <- 
  pop_id %>%
  select(min_year, max_year) %>%
  split(., pop_id$ID)

n0_nt <- 
  lapply(n0_nt, function(x) {
  tibble(year = seq(from = x$min_year, to = x$max_year, by = 1)) } ) %>%
  bind_rows(., .id = "ID") %>%
  mutate(ID = as.numeric(ID),
         year = as.character(year))

# join these data to the lpi pop data
n0_nt_dat <- 
  inner_join(n0_nt, lpi_pop, by = c("ID", "year"))

# in the LPI calculation, you need two consecutive years
# remove the first year for each of these time-series

n0_nt_dat <- 
  n0_nt_dat %>%
  group_by(ID) %>%
  filter(year != min(year)) %>%
  ungroup()

n0_nt_dat

# create a species x year matrix (columns = species, rows = years)
spp_year <- 
  split(n0_nt_dat, n0_nt_dat$year) %>%
  lapply(., function(x) {
    
    unique(x$Class)
    
  })

# create a vector of all species in the database
spp_list <- unique(n0_nt_dat$Class)

spp_year <- lapply(spp_year, function(x) { 
  
  tibble(species = spp_list,
         pa = if_else(spp_list %in% x, 1, 0))
  
  } ) %>%
  bind_rows(., .id = "year")

spp_year <- 
  spp_year %>%
  pivot_wider(names_from = species,
              values_from = pa)

# run an nmds on these data to see how the composition has changed through time
spp_year

# use multidimensional scaling (MDS) i.e. principle coordinates analysis
mds.1 <- cmdscale(vegdist(select(spp_year, -year), method = "jaccard"), 
                  k = 2, eig = T, add = T )

# calculate the variance explained by the mds.1
round(mds.1$eig*100/sum(mds.1$eig), 1)

# check the correlation between observed differences and actual dissimilarities
round(cor(vegdist(select(spp_year, -year), method = "jaccard"), 
          dist(mds.1$points), method = "spearman"), 3)

# create a dataframe for plotting
mds.1_raw <- 
  tibble(mds_1 = mds.1$points[, 1],
         mds_2 = mds.1$points[, 2]) %>%
  mutate(year = as.numeric(spp_year$year))


# analyse the number of time-series in different classes

prop_class <- 
  n0_nt_dat %>%
  group_by(year, Class) %>%
  summarise(n_class = n(), .groups = "drop") %>%
  group_by(year) %>%
  mutate(n_total = sum(n_class)) %>%
  ungroup() %>%
  mutate(class_proportion = n_class/n_total) %>%
  mutate(year = as.numeric(year))

# population sizes through time for all groups combined
pop_all <- 
  lpi_pop %>%
  filter(!is.na(population_size)) %>%
  mutate(population_size = log10(1 + population_size)) %>%
  group_by(year) %>%
  summarise(population_size_m = mean(population_size, na.rm = TRUE),
            population_size_sd = sd(population_size, na.rm = TRUE),
            n = n(),
            .groups = "drop") %>%
  mutate(population_size_se = (population_size_sd/sqrt(n)),
         year = as.numeric(year)) 

```



I had some time and energy so I looked into the LPI calculation in some more detail and also did some preliminary analysis on the LPI data to see how the input data has changed through time. Perhaps they correct for this in some way but the changes have been quite striking.


## How have the starting populations changed through time?

This is plot is just the mean (and SE) for all population values at each year in the dataset. There has been a decline through time which might indeed make the random fluctuations more meaningul. 

However, the complicated issue with this is these are just the actual population values. My reading of the code from rlpi package and the McCrae et al. (2017) paper is that they do a fair bit of extrapolation for time-series with incomplete data. For example, if the time-series has more than six population size values, say: 1971, 1974, 1975, 1976, 1980, 1982 and 1985. Population size for all years between 1971 and 1985 are interpolated using a generalised additive model. Therefore, for each pair of years (i.e. 1971-1972, 1972-1973 etc.) a d-value is calculated. When there are fewer than six population size values in a time-series they use a slighty different interpolation method but not for all years between the initial and final years in the time-series. 

Long story short is that plotted here are just the actual data but the LPI is calculated using interpolated population data as well which means that this is not completely reflective.

```{r, echo=FALSE}
ggplot(data = pop_all,
       mapping = aes(x = year, y = population_size_m)) +
  geom_errorbar(mapping = aes(ymin = population_size_m-population_size_se,
                              ymax = population_size_m+population_size_se),
                width = 0.1) +
  geom_point(alpha = 1, shape = 16, size = 2) +
  geom_smooth(size = 0.1, colour = "black") +
  scale_colour_viridis_d() +
  xlab("year") +
  ylab("log10(mean +- se population size)") +
  theme_classic()
```


## How has the population composition changed through time?

For this, I looked at the representation of different Classes (i.e. Aves, Amphibia) etc. through time. I'm more confident in this analysis because I included a species in the analysis for all years between the start and end years in the series. This partially circumvents the problem that plagues the analysis of changes in the population size through time.

There has been a clear shift in species composition through time. I showed this by using the Jaccard dissimilarity in Class-level composition through time using a PCoA. The y-axis is the first PCoA axis which explains 43% of the variation.

```{r, echo=FALSE}
ggplot(data = mds.1_raw,
       mapping = aes(x = year, y = mds_1)) +
  geom_jitter(size = 2.5, alpha = 0.9) +
  geom_smooth(size = 0.1, alpha = 0.5, colour = "black") +
  scale_colour_viridis_c() +
  ylab("MDS1 (43 %)") +
  theme_classic()
```

I then also looked at how the proportion of time-series in different classes has changed through time. Main patterns are a decline in proportion of mammal time-series and an increase in the proportion of Actinopterygii (ray-finned fishes). 

```{r, echo=FALSE}
ggplot(data = prop_class,
       mapping = aes(x = year, y = class_proportion, colour = Class)) +
  geom_point() +
  geom_smooth(method = "lm", size = 0.1, se = TRUE, alpha = 0.1) +
  scale_colour_viridis_d() +
  ylab("proportion time-series") +
  theme_classic()
```


## Conclusions

Main message from this is that if we want to see how the starting population sizes have changed through time and whether this affects the LPI, we will have to figure out how they do their interpolations. The rlpi code is not easy to follow but it would be doable to extract their code for how they do the interpolations with some serious effort.


## The end





